Estimating the number of clusters using Cross Validation

### 2015-08-22 cvclust.tex
Add proof and complete .tex file

Most existing methods are either model based method requires strong parameter
assumptions or ad hoc method utilizes within-cluster dispersion. Although many
view selecting the number of clusters as a model selection problem, very few
approaches this problem from the prediction point of view. Select model with
smallest prediction error via cross-validation is one of the simplest and most
widely used model selection techniques in supervised learning. The lack of
true class (label) in data set makes the adoption of cross-validation into
unsupervised leaning problem difficult. Perhaps the only exception is
\cite{tibshirani2005cluster}, which selects the optimal $k$ by prediction
strength. The strategy is to first cluster the test data and training data
into $k$ clusters respectively. Then, for each pair of observations that are
assigned to the same test cluster, algorithm determines whether they are also
assigned to the same cluster based on the training centers. The intuition here
is, if $k=k_0$, the true number of clusters, then the $k$ training set
clusters will be similar to the $k$ test clusters, and hence will predict them
well. However, a specifically defined prediction error measure is used in such
procedure, which is quite different from the one commonly used in
cross-validation procedure in supervised learning. Therefore the well
understood properties of cross-validation procedure in supervised learning
cannot be carried over to the prediction strength method, makes the
interpretation of its result difficult.

@@

\begin{figure}
\centering
\includegraphics[width=25pc]{demo/nullcorr/equal.pdf}
\caption{Cross-validation error on $10$ replicates, with the number of
clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
multivariate normal distribution with correlation $\rho$.  The Gabriel
cross-validation criterion chooses the correct answer $k = 1$ whenever
$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
\label{fig:nullcorr-equal}
\end{figure}

@@

\begin{proposition}
Suppose that $\{(X_i,Y_i)\}_{i=1}^{n+m}$ is data from a single fold of Gabriel
cross-validation, where each $(X,Y)$ pair in $\R^2$ is an independent draw
from an equiprobable mixture of two multivariate normal distributions with
identity covariance. Suppose that the first mixture component has mean $(\muX,
\muY)$, and the second has mean $(-\muX, -\muY)$, where $\muX > 0$ and
$\muY > 0$.  If [Wei, add condition here], then $\CV(2) < \CV(1)$ with
probability tending to one as $m$ and $n$ increase.
\end{proposition}

\begin{proof}
  [Wei, add proof here; you can take most of this from your document,
  ``LDA\_prior\_equal.tex'']
\end{proof}

We confirm the proposition with a simulation.

[Wei, describe simulation here, and include the figure]

@@

In this section, simulation is performed to evaluate the performance of our
proposed methods in locating the ``correct" number of clusters. We compare
with a basket of existing methods including Gap statistics
\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
\citep{fraley2002model}, CH-index \citep{calinski1974dendrite}, Hartigan
statistics \citep{hartigan1975clustering}, Jump method
\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
Bootstrap stability \citep{fang2012selection} in following simulation
settings. We select $L=R$ and $5-$fold cross-validation in row
($m=\frac{4}{5}n$) as default parameter setting for our proposed Gabriel
method. Note that set $L=R$ corresponding to $2-$fold cross-validation in
column.
\begin{enumerate}

@@

  \item Two clusters in $4$ dimensions --- $50$ i.i.d standard normal
    observations are generated, i.e. generated from multivariate normal
    $\mathcal{N}\left(\boldsymbol\mu,\mathbf{I}\right)$ for each cluster, with
    cluster center $\boldsymbol\mu = (1,0,0,1)$ and $(1,3.5,3.5,1)$
    respectively

@@

Besides the number of clusters selected by each algorithm is returned, we also
give the prediction error (PE) for each algorithm in each setting. The PE is
defined by 
\[
  \PE(k)
    =
      \frac{\|\hat{\mathbf{X}}_k- \boldsymbol{\mu} \|_F}
           {PE_{\text{oracle}}}
\]
where
\[
  \PE_{\text{oracle}} = \min_k \|\hat{\mathbf{X}}_k- \boldsymbol{\mu} \|_F
\]
Matrix $\boldsymbol{\mu}$ consists of true centers of observations in data
matrix $\mathbf{X}$, while $\hat{\mathbf{X}}_k$ is obtained by replacing each
observation in $\mathbf{X}$ by its closest cluster center result from applying
K-means on $\mathbf{X}$ with parameter $k$. Here $k$ is the optimal number of
clusters selected by each algorithm on data $\mathbf{X}$. This is only
possible on simulated data because otherwise one would never know the true
center matrix~$\boldsymbol{\mu}$.

The oracle selected number for each data $\mathbf{X}$ is also given on the
table, which is defined by 
\[
  K_{\text{oracle}}
    =
      \argmin_{k} \|\hat{\mathbf{X}}_k - \boldsymbol{\mu} \|_F
\]

\clearpage

\begin{center}
\footnotesize
\begin{longtable}{lcccccccccccccccc}
\caption{\label{table1} Simulation Results}\\
\toprule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
\midrule
\endfirsthead
\multicolumn{17}{c}%
  {\tablename\ \thetable\ -- \textit{Continued from previous page}} \\
\midrule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
\midrule
\endhead
\multicolumn{17}{r}{\textit{Continued on next page}} \\
\endfoot
\bottomrule
\multicolumn{17}{l}{\footnotesize
  \textsuperscript{$\ast$} CH, Hartigan, and Stability are excluded from
  Setting~1 because they cannot select $k=1$.
}
\endlastfoot
\input{demo/bench/setting1/results.tex}
\midrule
\input{demo/bench/setting2/results.tex}
\midrule
\input{demo/bench/setting3/results.tex}
\midrule
\input{demo/bench/setting4/results.tex}
\midrule
\input{demo/bench/setting5/results.tex}
\midrule
\input{demo/bench/setting6/results.tex}
\end{longtable}
\end{center}

@@

The ``\textendash'' sign in Setting~1 is because for CH-index, Hartigen and
Bootstrap stability, the minimum $k$ they can pick is $2$. So they are not
included for comparison since the true $k=1$. Setting $2-4$ have Gaussian
distribution, where the Gaussian mixture model-based clustering suppose to
perform the best. It is only the case for setting $2$. In setting $3$ and $4$,
it has a very poor performance even if the underlying distribution is indeed
Gaussian mixture model. It highlights the difficulties that high dimension
data could cause. Other methods (Gap statistics, CH-index, Hartigan, Jump
method, Prediction strength and Boostrap stability) also have problems in
these two high-dimension settings as we could not find any one with reasonable
performance in terms of finding the correct $k$ in both cases. In terms of PE,
Boostrap stability and Gap statistics have reasonable performance, but come
with the price of picking larger $k$. Our proposed Gabriel method (also Wold
CV method) clearly stands out from the basket of existing methods using either
measure under the high-dimension setting $3$ and $4$.

Setting $5$ and $6$ corresponding to the situation where data are heavy-tailed
instead of normal distributed, coupled with unequal number of observations or
unequal variance for different clusters. From Table \ref{table1}, we can see
that our proposed Gabriel method is robust and has superior performance
compare with other methods. Because one can hardly tell the underlying
distribution of data in practice, the resilience to non-Gaussian data gives
the Gabriel method a clear edge.

@@

The first one is congress voting data which consists of voting records of
$98$th United States Congress, $2$nd session. This data set includes votes for
each of the U.S. House of Representatives Congressmen on the $16$ key votes
identified by the \textit{CQA} (Congressional Quarterly Almanac). For each
vote, each Congressman either vote positively ``yea" (voted for/paired
for/announced for),  negatively ``nay'' (voted against/paired
against/announced against) or position unknown ``?". We took out those records
contain ``?" before comparing each algorithm. It results in $232$ remaining
records, with $124$ democrat and $108$ republican.

The second data set is the well-known Wisconsin breast cancer data set \citep{mangasarian1990pattern}. After excluding the records with missing data, this data set consists records of $683$ patients, each with measurements of nine attributes of their biopsy specimens. It is known that there exist at least two groups of patients: $444$ patients with benign specimens and $239$ patients with malignant specimens.\\

@@

\begin{table}[H]
\begin{center}
\caption{\label{table2} Number of clusters selected by each algorithm}
\begin{tabular}{lccccc}
    \hline                  
 & Congress Voting && Breast Cancer && Sonar \\ \hline                    
CH-index & $2$ && $2$  && $2$    \\
Hartigan & $3$ && $3$  &&  $3$   \\
Jump & $9$ && $9$ && $10$  \\   
Prediction strength & $2$ &&  $2$ && $1$    \\
Bootstrap stability & $2$ &&  $2$ &&  $10$   \\ 
Gap & $10$ && $9$ && $10$  \\   
Gaussian-Mix & $7$ && $5$ && $1$  \\   
Gabriel & $2$  && $3$  && $2$  \\    
Wold & $2$ && $3$ && $10$ \\  \hline 
\end{tabular}
\end{center}
\hspace{0.5in} \footnotesize {All the algorithms executed with their default parameter settings with $k$ ranges from $1$ to $10$}
\end{table}

Since most congressmen vote base on their parties' interest, $2$ parties
(Democratic and Republican) represent two clusters in this data set. So the
optimal number should be two. Close inspection shows \textit{kmeans} with
$k=2$ separates the two parties very well with the lowest miss-classification
error $10.43\%$ ($k=3$ has $14.78\%$). Note that CH-index and Bootstrap
stability also return $k=2$, but $2$ is the lower bound those methods can
select for $k$. So it's not clear they actually choose $k=2$ or they hit the
lower bound (they would pick $k=1$ if allowed). For the breast cancer data,
it's know to have at least $2$ cluster based on whether it's benign specimens
or not. But it doesn't mean the optimal $k$ should be $2$.
\cite{fujita2014non} noticed that the malign group is quite heterogeneous and
can be further clustered into at least two subgroups. Hence, the result $k=3$
given by our proposed Gabriel method (as well as Wold method) make sense. For
the Sonar data set which is relatively high-dimension, only our proposed
Gabriel method and CH-index selects $k=2$. Majority of the methods collapse to
either select the maximum or select the minimum number, underline how
difficult it was to pick the right $k$ when dimension increases.

>>>>

Most existing methods are either model based method requires strong parameter
assumptions or otherwise lack of clear interpretation and theoretical justification.
Although many view selecting the number of clusters as a model selection problem, very few
approaches this problem from the prediction point of view. Select model with
smallest prediction error via cross-validation is one of the simplest and most
widely used model selection techniques in supervised learning. The lack of
true class (label) in data set makes the adoption of cross-validation into
unsupervised leaning problem difficult. One exception is
\cite{tibshirani2005cluster}, which selects the optimal $k$ by prediction
strength. The strategy is to first cluster the test data and training data
into $k$ clusters respectively. Then, for each pair of observations that are
assigned to the same test cluster, algorithm determines whether they are also
assigned to the same cluster based on the training centers. The intuition here
is, if $k=k_0$, the true number of clusters, then the $k$ training set
clusters will be similar to the $k$ test clusters, and hence will predict them
well. However, a specifically defined prediction error measure is used in such
procedure, which is quite different from the one commonly used in
cross-validation procedure in supervised learning. Therefore the well
understood properties of cross-validation procedure in supervised learning
cannot be carried over to the prediction strength method, makes the
interpretation of its result difficult. \citet{wang2010consistent} also uses 
cross-validation to select the optimal $k$. Instead of selecting $k$ which minimizes 
prediction error, such method picks $k$ which minimizes the specifically defined clustering instability.

@@

\begin{figure}[H]
\centering
\includegraphics[width=25pc]{demo/nullcorr/equal.pdf}
\caption{Cross-validation error on $10$ replicates, with the number of
clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
multivariate normal distribution with correlation $\rho$.  The Gabriel
cross-validation criterion chooses the correct answer $k = 1$ whenever
$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
\label{fig:nullcorr-equal}
\end{figure}

@@

\begin{proposition}
Suppose that $\{(X_i,Y_i)\}_{i=1}^{n+m}$ is data from a single fold of Gabriel
cross-validation, where each $(X,Y)$ pair in $\R^2$ is an independent draw
from an equiprobable mixture of two multivariate normal distributions with
identity covariance. Suppose that the first mixture component has mean $(\muX,
\muY)$, and the second has mean $(-\muX, -\muY)$, where $\muX > 0$ and
$\muY > 0$.  If $1 + 2\Phi(\mu^Y)+ \frac{2\varphi(\mu^Y)}{\mu^Y} < 4\Phi(\mu^X)$, then $\CV(2) < \CV(1)$ with
probability tending to one as $m$ and $n$ increase.
\end{proposition}

\begin{proof}
There are two clusters $G_1$ and $G_2$, where observations from $G_1$ are distributed as
\[	\mathcal{N}\left( \begin{pmatrix} 
    \mu^X \\
    \mu^Y \\
  \end{pmatrix}, \mathbf{I} \right)	\]
and observations from $G_2$ are distributed as
\[	\mathcal{N}\left( \begin{pmatrix} 
    -\mu^X \\
    -\mu^Y \\
  \end{pmatrix}, \mathbf{I} \right)	\]
where $\mu^X_1 > 0$ and $\mu^Y_1 > 0$. Let $G_i$ be the true cluster where observation $i$ is generated from, by assumption
\[	P(G_i=G_1) = P(G_i=G_2) = 1/2	\]
After applying $k$-means on $\{ Y_i\}_{i=1}^{n}$ with $k=2$, if $n$ is large enough, we have the estimated centroids $\bar{\mu}^Y_1$ and $\bar{\mu}^Y_2$ be close to $E(Y \mid Y>0)$ and $\E(Y \mid Y < 0)$,
with errors will be of size $\OhP(n^{-1/2})$. Here
\begin{align}
E(Y \mid Y>0) &= E(Y_1 \mid Y_1 > 0) \cdot P(Y_1>0) + E(Y_2 \mid Y_2 > 0) \cdot P(Y_2>0) \nonumber \\ 
   &= 2\varphi(\mu^Y)+2\mu^Y\Phi(\mu^Y)-\mu^Y  \\  \nonumber 
\end{align}
where $Y_1 \sim N(\mu^Y, 1)$ and $Y_2 \sim N(-\mu^Y, 1)$, and we used Lemma~\ref{lem:truncated-normal-moments}
(Appendix~\ref{app:technical-lemmas}). Similarly, we have
\begin{align}
E(Y \mid Y<0) &= E(Y_1 \mid Y_1 < 0) \cdot P(Y_1<0) + E(Y_2 \mid Y_2 < 0) \cdot P(Y_2<0) \nonumber \\ 
   &= -2\varphi(\mu^Y)-2\mu^Y\Phi(\mu^Y)+\mu^Y  \\  \nonumber 
\end{align}
 where $\varphi()$ and $\Phi()$ are the standard normal probability and cumulative distribution function respectively.

Same as in single cluster case, the classification rule learned from
$\{(X_i, \hGY_i)\}_{i=1}^{n}$ variables will be determined according to
whether $X > 0$, with the decision boundary be at $0 + \OhP(n^{-1/2})$.
By symmetry, the CV error for points from $G_1$ is same as the points from $G_2$. Because $P(G=G_1) = P(G=G_2) =1/2$, the CV error can be calculated solely from $G_2$, that is

\begin{align*}
  \CV(2)
  &=
    \frac{1}{m}
    \sum_{i=n+1}^{n+m}
      \| (Y_i - \bmuY_1) 1\{\hGX_i = 1\} \|^2
      +
      \| (Y_i - \bmuY_2) 1\{\hGX_i = 2\} \|^2, \hspace{0.2in}Y\sim N(-\mu^Y,1)
\\
  &=
    \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]
      + \OhP(m^{-1/2}) + \OhP(n^{-1/2})\\
   \end{align*}
   With 
   \begin{align}
 \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]  =& P(\hGX_i = 1)\cdot E[(Y - a)^2] + P(\hGX_i = 2)\cdot E[(Y + a)^2] \nonumber \\  \nonumber
 =& [1-\Phi(\mu^X)][var(Y)+(-\mu^Y - a)^2]+\\ \nonumber
  & \Phi(\mu^X)[var(Y)+(-\mu^Y + a)^2]\\  \nonumber
  =& [1-\Phi(\mu^X)][1+(\mu^Y + a)^2]+\Phi(\mu^X)[1+(\mu^Y -a )^2]   \\  \nonumber
  =& 1+(\mu^Y + a)^2 -4a\Phi(\mu^X)\mu^Y  \nonumber
\end{align}
where $a$ is given by $(1)$.

When $k=1$, the result is straight forward since the estimated centroid will approximately equal to
$0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error
will be
\[
  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
         = 1 + (\mu^Y)^2 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
\]
So if we have $1 + 2\Phi(\mu^Y)+ \frac{2\varphi(\mu^Y)}{\mu^Y} < 4\Phi(\mu^X)$, we have $\CV(2) < \CV(1)$
\end{proof}

We confirm this result with a simulation.  We perform $10$ replicates for each pair of $(\mu^X, \mu^Y)$, where both $\mu^X$ and $\mu^Y$ take value on grid of $[0,3]$ with step $0.1$.  In each
replicate, we generate $20000$ observations from two multivariate normal distributions with identity covariance, where one has mean $(\muX, \muY)$ and the other one has mean $(-\muX, -\muY)$.
We perform a single $2 \times 2$ fold of Gabriel cross-validation and report the
times (out of $10$ replicates) when $k=2$ is selected by the algorithm in stead of $k=1$. Figure~\ref{fig:overlap-color_plot} shows the frequency $k=2$ is selected by the algorithm for each pair of $(\mu^X, \mu^Y)$. The red spot means high number (close to $10$) is selected by the algorithm, which means algorithm very likely will pick $k=2$ over $k=1$ for the corresponding $(\mu^X, \mu^Y)$. While blue spot means algorithm prefer $k=1$ for the corresponding value of $(\mu^X, \mu^Y)$. We can see the simulation result perfectly align with the theoretical curve (the black line), which separates the $k=2$ zone from the $k=1$ zone. It demonstrates that the Gabriel cross-validation works exactly as it suppose to under such setting. The position of red spots shows that when the two clusters are reasonably apart (not overlapping too heavily) in both dimensions, the Gabriel cross-validation is asymptotically consistent.

\begin{figure}[H]
\centering
\includegraphics[width=25pc]{demo/overlap/color_plot.pdf}
\caption{Number of times $k=2$ is selected out of $10$ replicates for each pair of $(\muX, \muY)$. The heat map shows the frequency $k=2$ is selected by the algorithm, with blue means low number (of $k=2$) is selected and red color means high number is selected. The black line is the theoretical curve, above which the algorithm suppose to pick $k=2$ and below which algorithm select $k=1$.  }
\label{fig:overlap-color_plot}
\end{figure}

@@

In this section, simulation is performed to evaluate the performance of our
proposed methods in locating the ``correct" number of clusters. We compare
with a basket of existing methods including Gap statistics
\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
\citep{fraley2002model}, CH-index \citep{calinski1974dendrite}, Hartigan
statistics \citep{hartigan1975clustering}, Jump method
\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
Bootstrap stability \citep{fang2012selection} in following simulation
settings. We select $p=q$ and $5-$fold cross-validation in row
($m=\frac{1}{4}n$) as default parameter setting for our proposed Gabriel
method. Note that set $p=q$ corresponding to $2-$fold cross-validation in
column.
\begin{enumerate}

@@

  \item Two clusters in $4$ dimensions --- $50$ i.i.d
    observations are generated from both multivariate normal
    $\mathcal{N}\left(\mu_1,0.5\Sigma\right)$ and multivariate normal
    $\mathcal{N}\left(\mu_2,1.5\Sigma\right)$, where 
    cluster center $\mu_1 = (1,0,0,1)$ and $\mu_2 =(1,3.5,3.5,1)$
    respectively. $\Sigma$ has $AR(1)$ structure with $\rho=-0.2$ and $\sigma=1$.

@@

Table \ref{table1} shows the distribution of $k$ selected by each algorithm in 
each simulation setting. We run $k$ from $1$ to $15$, and the best $k$ within this range
is selected by the algorithms. [ NA in the table means...]
\clearpage

\begin{center}
\footnotesize
\begin{longtable}{lcccccccccccccccc}
\captionsetup{justification=centering}
\caption{\label{table1} Simulation Results}\\
\toprule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
\midrule
\endfirsthead
\multicolumn{17}{c}%
  {\tablename\ \thetable\ -- \textit{Continued from previous page}} \\
\midrule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
\midrule
\endhead
\multicolumn{17}{r}{\textit{Continued on next page}} \\
\endfoot
\bottomrule
\multicolumn{17}{l}{\footnotesize
  \textsuperscript{$\ast$} CH, Hartigan, and Stability are excluded from
  Setting~1 because they cannot select $k=1$.
}
\endlastfoot
\input{demo/bench/setting1/results.tex}
\midrule
\input{demo/bench/setting2/results.tex}
\midrule
\input{demo/bench/setting3/results.tex}
\midrule
\input{demo/bench/setting4/results.tex}
\midrule
\input{demo/bench/setting5/results.tex}
\midrule
\input{demo/bench/setting6/results.tex}
\end{longtable}
\end{center}

@@

For CH-index, Hartigen and Bootstrap stability, the minimum $k$ they can pick is $2$. So they are not
included for comparison since the true $k=1$. It also demonstrates the difficulty 
that null scenario poses for some existing algorithms. For those could select $k=1$,
most of them work well under such setting except for Jump method. Setting $2-4$ have Gaussian
distribution, where the Gaussian mixture model-based method suppose to
perform the best. However, it only performs reasonably well in setting $2$. In setting $3$ and $4$,
it has a very poor performance even if the underlying distribution is indeed
Gaussian mixture model. It highlights the difficulty that high dimension
data could cause. Other methods (CH-index, Hartigan, Jump
method, Prediction strength and Boostrap stability) also have problems in
these two high-dimension settings as we could not find any one with reasonable
performance in terms of finding the correct $k$ in both cases. The only exception is
Gap statistics, which works perfectly in both settings. Our proposed Gabriel method (also Wold
CV method) clearly stands out from the basket of existing methods under the 
high-dimension setting $3$ and $4$, as we can see they both perform perfectly under
such settings.

Setting $5$ and $6$ corresponding to the situation where data are heavy-tailed
instead of normal distributed, coupled with unequal number of observations or
unequal variance for different clusters. From Table \ref{table1}, we can see none of 
the existing methods has good performance in both settings. In contrast,
our proposed Gabriel method is robust and has superior performance
compare with other methods. Because one can hardly tell the underlying
distribution of data in practice, the resilience to non-Gaussian data gives
the Gabriel method a clear edge.

@@

The first one is congress voting data which consists of voting records of
$98$th United States Congress, $2$nd session \citep{schlimmer1987concept}. 
This data set includes votes for each of the U.S. House of Representatives Congressmen on the $16$ key votes
identified by the \textit{CQA} (Congressional Quarterly Almanac). For each
vote, each Congressman either vote positively ``yea" (voted for/paired
for/announced for),  negatively ``nay'' (voted against/paired
against/announced against) or position unknown ``?". We took out those records
contain ``?" before comparing each algorithm. It results in $232$ remaining
records, with $124$ democrat and $108$ republican.

The second data set is the well-known Wisconsin breast cancer data set \citep{mangasarian1990pattern}. After excluding the records with missing data, this data set consists records of $683$ patients, each with measurements of nine attributes of their biopsy specimens. It is known that there exist at least two groups of patients: $444$ patients with benign specimens and $239$ patients with malignant specimens.

@@

\begin{table}[H]
\begin{center}
\captionsetup{justification=centering}
\caption{\label{table2} Number of clusters selected by each algorithm}
\begin{tabular}{lccccc}
    \hline                  
 & Congress Voting && Breast Cancer && Sonar \\ \hline                    
CH-index & $2$ && $2$  && $2$    \\
Hartigan & $3$ && $3$  &&  $3$   \\
Jump & $9$ && $9$ && $10$  \\   
Prediction strength & $2$ &&  $2$ && $1$    \\
Bootstrap stability & $2$ &&  $2$ &&  $10$   \\ 
Gap & $10$ && $9$ && $10$  \\   
Gaussian-Mix & $7$ && $5$ && $1$  \\   
Gabriel & $2$  && $3$  && $2$  \\    
Wold & $2$ && $3$ && $10$ \\  \hline 
\end{tabular}
\end{center}
\hspace{0.5in} \footnotesize {All the algorithms executed with their default parameter settings with $k$ ranges from $1$ to $10$}
\end{table}

Since most congressmen vote base on their parties' interest, $2$ parties
(Democratic and Republican) represent two clusters in this data set. So the
optimal number should be two. Close inspection shows $k$-means with
$k=2$ separates the two parties very well with the lowest miss-classification
error $10.43\%$ ($k=3$ has $14.78\%$). Note that CH-index and Bootstrap
stability also return $k=2$, but $2$ is the lower bound those methods can
select for $k$. So it's not clear they actually choose $k=2$ or they hit the
lower bound (they would pick $k=1$ if allowed). For the breast cancer data,
it's known to have at least $2$ cluster based on whether it's benign specimens
or not. But it doesn't mean the optimal $k$ should be $2$.
\cite{fujita2014non} noticed that the malign group is quite heterogeneous and
can be further clustered into at least two subgroups. Hence, the result $k=3$
given by our proposed Gabriel method (as well as Wold method) make sense. For
the Sonar data set which is relatively high-dimension, only our proposed
Gabriel method and CH-index selects $k=2$. Majority of the methods collapse to
either select the maximum or select the minimum number, underline how
difficult it was to pick the right $k$ when dimension increases.
